FlowFPX: Nimble Tools for Debugging Floating-Point Exceptions

### 2024-08-01 paper.tex
jcon2: editos

Reliable numeric computations are central to high-performance computing,
machine learning, and scientific applications.
Yet the \fp{} arithmetic that powers these computations is fundamentally
unreliable~(\cref{s:background}).
Exceptional values, such as Not a Number (\Nan{}) and infinity (\Inf{}),
can and often do arise thanks to culprits such as roundoff error,
catastrophic cancellation, singular matrices, and vanishing
derivatives~\cite{sdjmrstp-pc-2022,ddghlllprr-correctness-2022,gllprt-correctness-2021,fpchecker-reports,llg-soap-2022,bllmg-xloop-2022}.
Developers are responsible for guarding against exceptions, but this task
is difficult because many operations can generate and propagate exceptions.
Worst of all, operations such as less-than (\code{<}) can kill an exceptional value,
leaving no trace of the problem.
There is little tool support to assist in exception debugging,
thus (unsurprisingly) a quick GitHub search reports over 4,000 open issues
related to \NaN{} or \Inf{} exceptions~\cite{github-issues}.

This paper introduces \FlowFPX{}, a toolkit for debugging
\fp{} exceptions~(\cref{s:flowfpx})
that has helped improve a variety of applications,
from ocean simulations to heat modes~(\cref{s:casestudies}).
\FlowFPX{} consists of three tools that can work together to debug \fp{} computations:
\begin{enumerate}
  \item
    The centerpiece of \FlowFPX{} is \TF{}, a dynamic analysis tool that
    selectively monitors for exceptions and fuzzes Julia code for vulnerabilities.
    \TF{} is a Julia library that operates on Julia code.
  \item
    To visualize results, \FlowFPX{} uses coalesced stack trace graphs (CSTGs, or
    \CSTG{}s)~\cite{humphreySystematicDebuggingMethods2014}
    to summarize the program contexts that handled exceptional values.
    \CSTG{} is a generic stack trace graph summarization tool.
  \item
    A companion tool \GPUFPX{}~\cite{llsflg-hpdc-2023} provides fine-grained insights for GPU kernels.
    \GPUFPX{} is a dynamic library that programmers of GPU kernels can use.
\end{enumerate}

@@

The flip side is that most real numbers fall into the gaps between \fp{} numbers and must be rounded, which introduces error
and can lead to surprising results.
For example, adding the tiny Planck constant to the large Avogadro number results
in Avogadro's number after rounding:

@@

IEEE~754 defers the question of how to handle such exceptions to application code.
It is up to developers to watch for \Nan{}, \Inf{}, and subnormal numbers (underflow)
and implement an appropriate repair.
This is no easy task, however, because of the approximations inherent to
floating point.
Even an apparently-safe division could result in a \Nan{} if its
denominator gets truncated to zero.
Some \Nan{}s might be spurious, others might be fatal, but in any event
anticipating the various exceptions is a burden.
All too often, exceptional values go unhandled and flow through the code.

@@

\FlowFPX{} (FPX: Floating Point eXeption) is a toolkit for tracking down \fp{} exceptions.
The primary tools in the \FlowFPX{} toolkit are \TF{}, which records lifetimes
and enables fuzzing, and \CSTG{}s (more precisely, CSTGs), which visualize
the flow of exceptions.
\GPUFPX{} is a third component of \FlowFPX{} that tracks \fp{} exceptions inside GPU kernels~\cite{llsflg-hpdc-2023}.
This section covers each tool in detail.

@@

\TF{} tracks exceptional values in Julia programs across
their \genpropkill{} lifetime and can fuzz code for vulnerabilities by
injecting a \Nan{} or \Inf{} as the result of an operation. \TF{} is
implemented in Julia as a library and is available on JuliaHub:

@@

Crucially, the tracked version of an exception, say a \NaN{}, points
to the original exceptional value.
This enables techniques such as \NaN{}-packing, or otherwise
using the payload in creative ways.
Not all Julia operations currently preserve \NaN{}s
bit-for-bit,\footnote{\url{https://github.com/JuliaLang/julia/issues/48523}}
but \TF{} does its part.

@@

In addition to enabling tracking exceptional values, operator
overloading gives \TF{} a powerful way to fuzz code from the inside
out.
Each overloaded function serves as a hook where \TF{} can decide whether to observe
the operation or step in, discard the original result, and return an exception instead.
Injecting faults in a random way~\cite{hamlet1994random}, also known as fuzzing,
is a useful way to discover vulnerabilities in a large codebase.
Demmel et~al. propose essentially the same idea for BLAS and
LAPACK~\cite{ddghlllprr-correctness-2022}.

When fuzzing is enabled, every time \TF{} intercepts a \fp{}
operation, it decides whether or not it should return a \Nan{} instead
of the actual computed value.
This decision is based on several parameters that the developer can
tune to control fuzzing:

\begin{itemize}
\item \texttt{active::Bool} fuzz only when set to true.
\item \texttt{odds::Int64} inject if \texttt{rand(odds) == 1}.
\item \texttt{n\_inject::Int64} upper bound on the number of \Nan{}s to inject.
\item \texttt{functions::Array{String}} limit fuzzing to the dynamic extent of the listed functions.
\item \texttt{libraries::Array{String}} limit fuzzing to functions from the following libraries
\end{itemize}

\TF{} looks at these parameters in conjunction with stack trace at the
point where the operation was intercepted to make the decision.
When fuzzing is active (\texttt{active == true}), there are \Nan{}s
remaining to be injected (\texttt{n\_inject > 0}), and
\texttt{rand(odds)} returns \texttt{1}, then \TF{} inspects the
current stack trace to see if the point is in the dynamic extend of
the wanted \texttt{functions} or \texttt{libraries}.
When all the conditions are met, the overloaded operator returns
\Nan{} instead of whatever value would have been computed normally.
The value \texttt{n\_inject} is decremented before the next operation
gets intercepted.

Since \TF{} checks if it should inject a \Nan{} at \emph{every} \fp{}
operation it intercepts, means that \Nan{}s can be injected deep
inside of an algorithm instead of just at the inputs.
This can reveal vulnerabilities that otherwise would have been
difficult to catch with input-based fuzzers.

Developers can use the \texttt{functions} and \texttt{libraries}
parameters to narrow down the region where \Nan{}s can be injected.
It might not be useful, for instance, to fuzz test code that a
developer has no control over; a developer may not, for example, wish
to fuzz the standard library and instead focus on the code that they
are developing.

@@

\subsubsection{\TF{} Internals}

@@

Every arithmetic operation requires a similar overloading.
Thus, the implementation of \TF{} uses a metaprogramming technique
adapted from \texttt{Sherlogs}~\cite{kMilanklSherlogsJl2021}
to abstract over the common patterns.
For every binary operation, the library creates an overloading similar
to the one for \code{+} above.
Unary operations and others work analogously.
This approach saved thousands of lines of code.
The implementation weighs in at 218 lines and defines 645 overloaded function;
assuming 5 lines of code per function, a handwritten version would require over 3,000 lines.

@@

\TF{} can produce copious amounts of log files which can be challenging to sift through manually.
Coalesced stack trace graphs (CSTGs or \CSTG{}s for short) provide a way to visualize large amounts of stack traces in a compact form~\cite{humphreySystematicDebuggingMethods2014}.
This technique pairs well with \TF{} and makes analyzing the log files easier.

@@

Many programs offload work onto GPUs, which are no less susceptible to \fp{} exceptions than CPUs.
In fact, GPU programs are worse off because they lack exception-handling mechanisms from the CPU world~\cite{llg-soap-2022}.
Since \TF{} instruments Julia programs, it cannot help directly;
however, the companion tool \GPUFPX{} instruments GPU kernels to detect and report
\fp{} exceptions~\cite{llsflg-hpdc-2023}.
Together, \TF{} and \GPUFPX{} provide insights for accelerated programs.

\TF{} and \GPUFPX{} have no formal connection---that is, one does not
need to use \GPUFPX{} with \TF{}, nor vice-versa.
These two tools compliment one another in debugging mixed-paradigm
workloads, but they can be used in isolation.
\GPUFPX{} is a library exclusively for NVIDIA CUDA code and has no
special relation to Julia.
Nevertheless, in cases where Julia code leverages a CUDA kernel,
\GPUFPX and \TF{} can both aid in tracking down exceptional \fp{}
values.

@@

We can see that the \NaN{} appeared as the result of subtracting two infinities (\code{-Inf - -Inf}).
The trace further shows that this subtraction happens inside the function \texttt{momentum\_u!} on line 246 of the \texttt{rhs.jl} file.

This solves the mystery of where one \NaN{} came from, but raises a new question
about the source of the \Inf{} value.
The logs for \Inf{} gens have an answer:

\begin{lstlisting}
^([-1.515f31, 2])          TF/TrackedFloat.jl:138
literal_pow()              intfuncs.jl:325
...
materialize(^)             broadcast.jl:860
top-level getproperty(...) examples/sw_nan_tf.jl:14
..
\end{lstlisting}
%
This \Inf{} came from an exponent that overflowed the float type (\code{-1.515e31\string^2}).
\TF{} has shown exactly which numeric values in which operation
caused exceptions to occur.

@@

\Cref{fig:sw_nan_cstg} presents the stack graph for \NaN{} gen events
in \ShallowWaters{} with the high CFL number.
Reading bottom-up, every \NaN{} came from calls to the `\code{-}' and `\code{+}' operations.
Calls to `\code{+}' account for most of the \NaN{}s.
These \NaN{}s arose in two different contexts: a small number (30)
occurred within a momentum calculation, while the rest (162)
occurred within a continuity step.
Moving to the top of the graph, it shows that the function \code{run\_model} drove
the entire simulation.
With this overview of the program, a promising next step is to guard against \NaN{}s
in the momentum and continuity functions.

@@

Demmel et al.\cite{ddghlllprr-correctness-2022} examine \fp{} exceptional value handling in BLAS and LAPACK, identify several inconsistencies, and propose an API for debugging and adding determinism.
The proposal includes an extension to the \texttt{INFO\_ARRAY} parameter with fields that record \genpropkill{} information.
It also includes fuzzing, which we realize in \TF{}.

@@

Herbie~\cite{panchekhaAutomaticallyImprovingAccuracy2015} automatically rewrites arithmetic expressions to reduce \fp{} error.
This would combine well with \TF{}: first identify the location of a \NaN{}, ask Herbie to find a repair.
The Odyssey\cite{misbackOdysseyInteractiveWorkbench2023} workbench provides an interactive interface to Herbie.

>>>>

Reliable numeric computations are central to high-performance computing,
machine learning, and scientific applications.
Yet the \fp{} arithmetic that powers these computations is fundamentally
unreliable~(\cref{s:background}).
Exceptional values, such as Not a Number (\Nan{}) and infinity (\Inf{}),
can and often do arise thanks to culprits such as roundoff error,
catastrophic cancellation, singular matrices, and vanishing
derivatives~\cite{bllmg-xloop-2022,ddghlllprr-correctness-2022,fpchecker-reports,gllprt-correctness-2021,llg-soap-2022,sdjmrstp-pc-2022}.
Developers are responsible for guarding against exceptions, but this task
is difficult because many operations can generate and propagate exceptions.
Worst of all, operations such as less-than (\code{<}) can kill an exceptional value,
leaving no trace of the problem.
There is little tool support to assist in exception debugging,
thus (unsurprisingly) a quick GitHub search reports over 4,000 open issues
related to \NaN{} or \Inf{} exceptions~\cite{github-issues}.

This paper introduces \FlowFPX{}, a toolkit for debugging
\fp{} exceptions~(\cref{s:flowfpx})
that has helped improve a variety of applications,
from ocean simulations to heat modes~(\cref{s:casestudies}).
\FlowFPX{} consists of three tools that can work together to debug \fp{} computations:
\begin{enumerate}
  \item
    The centerpiece of \FlowFPX{} is \TF{}, a dynamic analysis tool written in Julia
    for Julia code that monitors exceptions and fuzzes for vulnerabilities.
  \item
    To visualize results, \FlowFPX{} uses coalesced stack trace graphs (CSTGs, or
    \CSTG{}s)~\cite{humphreySystematicDebuggingMethods2014}.
    Our CSTG library is written in \CPP{} and accepts text-format stack traces as input.
  \item
    For programs that offload work to CUDA GPUs,
    the binary instrumentation tool \GPUFPX{}~\cite{llsflg-hpdc-2023}
    adds logging to kernels.
\end{enumerate}

@@

The flip side is that most real numbers fall into the gaps between \fp{} numbers and must be rounded, which introduces error
and can lead to surprising results.
For example, adding the tiny Planck constant to the large Avogadro number results in
the Avogadro number after rounding:

@@

IEEE~754 defers the question of how to handle such exceptions to application code.
It is up to developers to watch for \Nan{}, \Inf{}, and subnormal numbers (underflow)
and implement an appropriate repair.
This is a difficult task because of the approximations inherent to
floating point.
Even an apparently-safe division could result in a \Nan{} if its
denominator gets truncated to zero.
Some \Nan{}s might be spurious while others might be fatal, but in any event
anticipating the various exceptions is a burden.
All too often, exceptional values go unhandled and flow through the code.

@@

\FlowFPX{} (FPX: Floating Point eXception) is a toolkit for tracking down \fp{} exceptions.
The primary tools in the \FlowFPX{} toolkit are \TF{}, which records lifetimes
and enables fuzzing, and \CSTG{}s (more precisely, CSTGs), which visualize
the flow of exceptions.
\GPUFPX{} is a third component of \FlowFPX{} that tracks \fp{} exceptions inside GPU kernels~\cite{llsflg-hpdc-2023}.
This section covers each tool in detail.

@@

\TF{} tracks exceptional values in Julia programs across
their \genpropkill{} lifetime and can fuzz code for vulnerabilities by
injecting a \Nan{} or \Inf{} as the result of an operation. \TF{} is
implemented in Julia as a library and is available on JuliaHub (TODO update link):

@@

Crucially,
\TF{} does nothing to modify the bit-level representation of floats or exceptions.
The tracked version of, say, a \NaN{}, points to the original exceptional
value.
Thus \NaN{}-packing and other creative uses of the payload continue to
work---provided the code does not use some other Julia operation
that does not preserve \NaN{}s bit-for-bit.\footnote{\url{https://github.com/JuliaLang/julia/issues/48523}}

@@

Julia's operator overloading lets us fuzz code from the inside
out.
Each overloaded function serves as a hook where \TF{} can decide whether to observe
the operation or replace the original result with an exceptional value.
Injecting faults in a random way~\cite{hamlet1994random}, also known as fuzzing,
can discover vulnerabilities in a large codebase.
Demmel et~al. propose essentially the same idea for BLAS and
LAPACK~\cite{ddghlllprr-correctness-2022}.

When fuzzing is enabled, every time \TF{} intercepts a \fp{}
operation, it decides whether or not it should return a \Nan{} instead
of the actual computed value.
This decision is based on several parameters:

\begin{itemize}
\item $\,$ \texttt{active::Bool} fuzz only when set to true.
\item $\,$ \texttt{odds::Int64} inject at each candidate location with probability $1 / \texttt{odds}$.
\item $\,$ \texttt{n\_inject::Int64} upper bound on the number of \Nan{}s to inject.
\item $\,$ \texttt{functions::Array{String}} limit fuzzing to the dynamic extent of the listed functions.
\item $\,$ \texttt{libraries::Array{String}} limit fuzzing to functions from the following libraries
\end{itemize}

\TF{} looks at these parameters in conjunction with stack trace at the
point where the operation was intercepted to make the decision.
When fuzzing is active (\texttt{active == true})
and there are \Nan{}s remaining to be injected (\texttt{n\_inject > 0})
and \texttt{rand(1:odds)} returns \texttt{1}, then \TF{} inspects the
current stack trace to see if the point is in the dynamic extent of
the declared \texttt{functions} or \texttt{libraries} (if any).
When all these conditions are met,
\TF{} returns a \NaN{} instead of calling the overloaded operation.
The value \texttt{n\_inject} is decremented before the next operation
gets intercepted.

Every intercepted operation is a candidate for injection,
which means that injected \Nan{}s can appear deep
inside of an algorithm instead of merely at its toplevel inputs.
This can reveal vulnerabilities that are difficult to catch
with input-based testing.
The parameters \texttt{functions} and \texttt{libraries} let
users control how deep to fuzz; for example, it is often helpful
to test specific libraries and avoid trusted standard ones.

@@

\subsubsection{Internals}

@@

Every arithmetic operation requires a similar overloading.
The implementation of \TF{} uses a metaprogramming technique
adapted from \texttt{Sherlogs}~\cite{kMilanklSherlogsJl2021}
to abstract over the common patterns.
For every binary operation, the library creates an overloading similar
to the one for \code{+} above.
Unary operations and others work analogously.
This approach saved thousands of lines of code.
The implementation weighs in at 218 lines and defines 645 overloaded function;
assuming 5 lines of code per function, a handwritten version would require over 3,000 lines.

@@

\TF{} can produce copious amounts of log files which can be challenging to sift through manually.
Coalesced stack trace graphs (CSTGs or \CSTG{}s for short) provide a way to visualize large amounts of stack traces in a compact form~\cite{humphreySystematicDebuggingMethods2014}.
This technique pairs well with \TF{} and eases the task of analyzing log files.

@@

Many programs offload work onto GPUs, which are no less susceptible to \fp{} exceptions than CPUs.
In fact, GPU programs are worse off because they lack exception-handling mechanisms from the CPU world~\cite{llg-soap-2022}.
Since \TF{} instruments Julia programs, it cannot help directly;
however, the companion tool \GPUFPX{} instruments GPU kernels
(specifically CUDA) to detect and report
\fp{} exceptions~\cite{llsflg-hpdc-2023}.
There is no formal connection between \TF{} and \GPUFPX{}.
Developers who choose to apply both tools to the respective Julia and CUDA
components of their code can gain insights for accelerated programs.

@@

We can see that the \NaN{} appeared as the result of subtracting two infinities (\code{-Inf - -Inf}).
The trace further shows that this subtraction happens inside the function \texttt{momentum\_u!} on line 246 of the \texttt{rhs.jl} file.
This solves the mystery of where one \NaN{} came from, but raises a new question
about the source of the \Inf{} value.
The logs for \Inf{} gens have an answer:

\begin{lstlisting}
^([-1.515f31, 2])          TF/TrackedFloat.jl:138
literal_pow()              intfuncs.jl:325
...
materialize(^)             broadcast.jl:860
top-level getproperty(...) examples/sw_nan_tf.jl:14
\end{lstlisting}
%
This \Inf{} came from an exponent that overflowed the float type (\code{-1.515e31\string^2}).
\TF{} has shown exactly which numeric values in which operation
caused exceptions to occur.

@@

\Cref{fig:sw_nan_cstg} presents the stack graph for \NaN{} gen events
in \ShallowWaters{} with the high CFL number.
Reading bottom-up, every \NaN{} came from calls to the `\code{-}' and `\code{+}' operations.
Calls to `\code{+}' account for most of the \NaN{}s.
These \NaN{}s arose in two different contexts:
a momentum calculation (30~\NaN{}s) and
a continuity step (162~\NaN{}s).
Moving to the top of the graph, it shows that the function \code{run\_model} drove
the entire simulation.
With this overview of the program, a promising next step is to guard against \NaN{}s
in the momentum and continuity functions.

@@

Demmel et~al.~\cite{ddghlllprr-correctness-2022} examine \fp{} exceptional value handling in BLAS and LAPACK, identify several inconsistencies, and propose an API for debugging and adding determinism.
The proposal includes an extension to the \texttt{INFO\_ARRAY} parameter with fields that record \genpropkill{} information.
It also includes fuzzing, which we realize in \TF{}.

@@

Herbie~\cite{panchekhaAutomaticallyImprovingAccuracy2015} automatically rewrites arithmetic expressions to reduce \fp{} error.
This would combine well with \TF{}: first identify the location of a \NaN{}, ask Herbie to find a repair.
The Odyssey~\cite{misbackOdysseyInteractiveWorkbench2023} workbench provides an interactive interface to Herbie.
